Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-11-03 manuscript/manuscript.Rmd
Added internal figure references.

```{=latex}
\begin{equation}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
One limitation of the Box-Cox transformation is that it is only defined for $x_i > 0$. In contrast, the Yeo-Johnson transformation under the transformation parameter $\lambda$ is defined for any $x_i \in \mathbb{R}$:

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is very sensitive to the presence of outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

@@

Box-Cox and Yeo-Johnson transformations are modified by adding a shift parameter $x_0$. The shift-sensitive Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$ and shift parameter $x_0$ is then:

```{=latex}
\begin{equation}
\phi_{\text{BC}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left(x_i - x_0 \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i - x_0) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x_i - x_0 > 0$. Likewise, the shift-sensitive Yeo-Johnson transformation of a feature value $x_i$ under transformation parameter $\lambda$ and shift parameter $x_0$ is:

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left( 1 + x_i - x_0\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log(1 + x_i - x_0) & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - x_i - x_0\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0< 0\\
-\log(1 - x_i - x_0) & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```
For both shift-sensitive transformations, $\lambda$ and $x_0$ are obtained by maximising the normal log-likelihood function. For the shift-sensitive Box-Cox transformation, and ignoring constant terms, this can be written as:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{BC}}, \hat{x}_0^{\text{BC}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \log(x_i - x_0)
\end{equation}
```
where $\sigma^2$ is the variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$. Similarly for the shift-sensitive Yeo-Johnson transformation:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{YJ}}, \hat{x}_0^{\text{YJ}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|)
\end{equation}
```
where $\sigma^2$ is the variance of Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (\mathbf{X})$.

@@

In essence, obtaining robust transformation parameters that provide central normality requires identifying outliers in the data and weighting such instances during the optimisation process. @Raymaekers2021-kq do this (during the final steps of their iterative procedure) through a weighted maximum likelihood estimation. This weighted maximum log-likelihood estimator is based on equations **INSERT**. Here, the estimator is further normalised by the total weight and extended to include the shift parameter. The shift-sensitive Box-Cox power transformation is:

@@

@@

.plot_weighting_functions(plot_theme = plot_theme)
```
Other criteria aside from MLE can be used for finding robust transformation parameters. We define robust criteria in Appendix A **INSERT**.

@@

The test statistic $\tau_{\text{gof}}$ is then defined as:
```{=latex}
\begin{equation}
\tau_{\text{gof}} = \frac{\sum_{i=1}^{N} r_i \left[r_i \in \mathbf{R}_{\text{central}}\right]}{\sum_{i=1}^N \left[r_i \in \mathbf{R}_{\text{central}}\right]} 
\end{equation}
```
Here $[\quad]$ denotes an Iverson bracket. The test statistic is equal to the mean of the residual errors of the central portion of the data.

@@

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to shifts, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure **INSERT**.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (Eq. **INSERT** and **INSERT**) and the shift-sensitive definitions (Eq. **INSERT** and **INSERT**) for each distribution, after adding a positive value $d$ to each distribution. Here, $d \in [1, 10^6]$.

@@

The optimal weighting function parameters for weighted MLE are shown in Tables **INSERT** and **INSERT** for robust, shift-sensitive Box-Cox and Yeo-Johnson transformations, respectively.

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Box-Cox} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Yeo-Johnson} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

Figure **INSERT** shows the distribution of errors $\left| \hat{\lambda}^r - \hat{\lambda}_0 \right|$ for non-robust transformations, as well as robust transformations using the optimal weighting function parameters. In the presence of outliers on either side of the distribution, the Yeo-Johnson transformation yields smaller errors than the Box-Cox transformation. For both transformations, weighting based on empirical probabilities yields the most consistent $\lambda$ parameter estimates in the presence of outliers. Weighting based on residual errors tended to perform least well, and were generally similar to (Yeo-Johnson) or worse than (Box-Cox) non-robust transformations.

```{r optimised_weighting_function_parameters, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\hat{\\lambda}^r$ ",
  "and expected $\\hat{\\lambda}_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the shift-sensitive Box-Cox transformation, ",
  "and shift-sensitive Yeo-Johnson transformation, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

@@

Figure *Insert reference* shows the residual errors of the transformed distribution as function of the empirical probability. As expected, the largest deviations from normality appear on the extremities of this range. For Box-Cox, the effect of outliers on the lower end of the distribution leads to noticeable asymmetry. Between very low and very high empirical probabilities, residual errors are constrained and relatively flat. This is the candidate range for the central portion of the data.

```{r goodness-of-fit-1, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, shift-sensitive transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, with 10 randomly drawn outliers distributions, ",
  "(up to 10% of samples) each. Larger errors occur for more extreme values at the edges of each distribution. ",
  "For comparison, the 99th percentile of residual errors for transformations of outlier-free distributions are shown (grey dashed line).")

@@

```
We then consider the empirical probability of type I errors: the probability of incorrectly classifying a centrally normal distribution as being non-normal based on the test statistic. Under the assumption that the asymmetric generalised normal distributions are centrally normal after robust transformation, this produces the relationship shown in Figure *Insert*. Since for $\kappa \leq 0.80$ error curves are similar, we fix the value of the central portion $\kappa$ to 0.80. This leads to test statistic values listed in Table *insert*, which derive from the type I error rates of the Yeo-Johnson transformation.

```{r goodness-of-fit-2, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic for the central portion $\\kappa$ of the distribution.")

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Probability of central normality, based on type I error rates for Box-Cox transformations.}
\begin{tabular}{c c c c c c c}

@@

In figure *insert* we apply the empirical goodness of fit test to assess central normality of features that are composed of a mixture of samples drawn from two normal distributions. With increased separation of the underlying normal distributions, the probability of the feature being centrally normal decreases, as expected.

```{r goodness-of-fit-3, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Bi-modal distributions and empirical goodness of fit test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical goodness of fit test to compute p-values for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

>>>>

```{=latex}
\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
One limitation of the Box-Cox transformation is that it is only defined for $x_i > 0$. In contrast, the Yeo-Johnson transformation under the transformation parameter $\lambda$ is defined for any $x_i \in \mathbb{R}$:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is very sensitive to the presence of outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing a shift parameter $x_0$ into equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The shift-sensitive Box-Cox transformation of a feature value $x_i$ of feature $\mathbf{X}$ under transformation parameter $\lambda$ and shift parameter $x_0$ is then:

```{=latex}
\begin{equation}
\label{eqn:box-cox-shift}
\phi_{\text{BC}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left(x_i - x_0 \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i - x_0) & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x_i - x_0 > 0$. Likewise, the shift-sensitive Yeo-Johnson transformation of a feature value $x_i$ under transformation parameter $\lambda$ and shift parameter $x_0$ is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift}
\phi_{\text{YJ}}^{\lambda, x_0} (x_i) = 
\begin{cases}
\left( \left( 1 + x_i - x_0\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log(1 + x_i - x_0) & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - x_i - x_0\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0< 0\\
-\log(1 - x_i - x_0) & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```
For both shift-sensitive transformations, $\lambda$ and $x_0$ are obtained by maximising the normal log-likelihood function. For the shift-sensitive Box-Cox transformation, and ignoring constant terms, this can be written as:

```{=latex}
\begin{equation}
\label{eqn:box-cox-shift-log-likelihood}
\left\{ \hat{\lambda}^{\text{BC}}, \hat{x}_0^{\text{BC}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \log(x_i - x_0)
\end{equation}
```
where $\sigma^2$ is the variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$. Similarly for the shift-sensitive Yeo-Johnson transformation:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift-log-likelihood}
\left\{ \hat{\lambda}^{\text{YJ}}, \hat{x}_0^{\text{YJ}} \right\} = \argmax_{\lambda, x_0} \sum_{i=1}^n -\frac{1}{2}\log(\sigma^2) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|)
\end{equation}
```
where $\sigma^2$ is the variance of Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (\mathbf{X})$.

@@

In essence, obtaining robust transformation parameters that provide central normality requires identifying outliers in the data and weighting such instances during the optimisation process. @Raymaekers2021-kq do this (during the final steps of their iterative procedure) through a weighted maximum likelihood estimation. This weighted maximum log-likelihood estimator is based on equations \ref{eqn:box-cox-shift-log-likelihood} and \ref{eqn:yeo-johnson-shift-log-likelihood}. Here, the estimator is further normalised by the total weight and extended to include the shift parameter. The shift-sensitive Box-Cox power transformation is:

@@

Other criteria aside from MLE can be used for finding robust transformation parameters, see the Appendix. For the remainder, we will use MLE as the optimisation criterion.

@@

.plot_weighting_functions(plot_theme = plot_theme)
```

@@

The test statistic $\tau_{\text{ecn}}$ is then defined as:
```{=latex}
\begin{equation}
\tau_{\text{ecn}} = \frac{\sum_{i=1}^{N} r_i \left[r_i \in \mathbf{R}_{\text{central}}\right]}{\sum_{i=1}^N \left[r_i \in \mathbf{R}_{\text{central}}\right]} 
\end{equation}
```
Here $[\quad]$ denotes an Iverson bracket. The test statistic is equal to the mean of the residual errors of the central portion of the data.

@@

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to shifts, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{box-cox-original} and \ref{yeo-johnson-original}) and the shift-sensitive definitions (equations \ref{box-cox-shift} and \ref{yeo-johnson-shift}) for each distribution, after adding a positive value $d$ to each distribution. Here, $d \in [1, 10^6]$.

@@

The optimal weighting function parameters for weighted MLE are shown in Tables \ref{tab:optimal-weighting-parameters-box-cox} and \ref{tab:optimal-weighting-parameters-yeo-johnson} for robust, shift-sensitive Box-Cox and Yeo-Johnson transformations, respectively.

```{=latex}
\begin{table}
\begin{center}
\label{tab:optimal-weighting-parameters-box-cox}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Box-Cox} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Yeo-Johnson} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

Figure \ref{fig:optimised-weighting-function-parameters} shows the distribution of errors $\left| \hat{\lambda}^r - \hat{\lambda}_0 \right|$ for non-robust transformations, as well as robust transformations using the optimal weighting function parameters. In the presence of outliers on either side of the distribution, the Yeo-Johnson transformation yields smaller errors than the Box-Cox transformation. For both transformations, weighting based on empirical probabilities yields the most consistent $\lambda$ parameter estimates in the presence of outliers. Weighting based on residual errors tended to perform least well, and were generally similar to (Yeo-Johnson) or worse than (Box-Cox) non-robust transformations.

```{r optimised-weighting-function-parameters, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{fig:optimised-weighting-function-parameters}",
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\hat{\\lambda}^r$ ",
  "and expected $\\hat{\\lambda}_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the shift-sensitive Box-Cox transformation, ",
  "and shift-sensitive Yeo-Johnson transformation, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

@@

Figure \ref{tab:empirical-central-normality-residual-error} shows the residual errors of the transformed distribution as function of the empirical probability. As expected, the largest deviations from normality appear on the extremities of this range. For Box-Cox, the effect of outliers on the lower end of the distribution leads to noticeable asymmetry. Between very low and very high empirical probabilities, residual errors are constrained and relatively flat. This is the candidate range for the central portion of the data.

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{tab:empirical-central-normality-residual-error}",
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, shift-sensitive transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, with 10 randomly drawn outliers distributions, ",
  "(up to 10% of samples) each. Larger errors occur for more extreme values at the edges of each distribution. ",
  "For comparison, the 99th percentile of residual errors for transformations of outlier-free distributions are shown (grey dashed line).")

@@

```
We then consider the empirical probability of type I errors: the probability of incorrectly classifying a centrally normal distribution as being non-normal based on the test statistic. Under the assumption that the asymmetric generalised normal distributions are centrally normal after robust transformation, this produces the relationship shown in Figure \ref{tab:empirical-central-normality-type-1-error-rate}. Since for $\kappa \leq 0.80$ error curves are similar, we fix the value of the central portion $\kappa$ to 0.80. This leads to test statistic values listed in Table \ref{tab:empirical-central-normality}, which derive from the type I error rates of the Yeo-Johnson transformation.

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{tab:empirical-central-normality-type-1-error-rate}",
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic for the central portion $\\kappa$ of the distribution.")

@@

```{=latex}
\begin{table}
\begin{center}
\label{tab:empirical-central-normality}
\caption{Probability of central normality, based on type I error rates for Box-Cox transformations.}
\begin{tabular}{c c c c c c c}

@@

In Figure \ref{fig:empirical-central-normality-examples} we apply the empirical central normality test to assess central normality of features that are composed of a mixture of samples drawn from two normal distributions. With increased separation of the underlying normal distributions, the probability of the feature being centrally normal decreases, as expected.

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{fig:empirical-central-normality-examples}",
  "Bi-modal distributions and empirical goodness of fit test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical goodness of fit test to compute p-values for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)
